Bayesian Reconstruction of Coevolutionary Histories

### 2013-09-25 paper/article.tex
General progress

Organismic symbioses, or interactions between individuals of two or more different species, constitute a fundamental aspect of many living systems and occur across the biological spectrum. The characteristics of symbiotic relationships can vary greatly, particularly in degrees of cooperation (parasitism vs. mutualism), fidelity (generalized symbionts utilizing multiple species vs. specialized, species-specific symbionts), and obligation (optional vs. vital to survival). Understanding how these relationships evolve and, in particular, how symbionts affect their partners' evolution---that is, the coevolutionary processes driven by symbiotic interactions---remain important questions in ecology and evolutionary biology.

@@

\textcite{Haffner:1988} were some of the first researchers to apply these new phylogenetic methods

@@

@@

\begin{figure}
\centering
\pscophylogeny
\caption{An example reconstruction of a coevolutionary history, where notation follows \Fref{fig:events}.}
\label{fig:cophylogeny}
\end{figure}

@@

Given data~$D = \left(d_H,d_S,d_R\right)$, the host sequence data, the symbiont sequence data, and the leaf--leaf associations between the host and symbiont trees, respectively, I define the posterior probability of the host tree~$H$ and the symbiont tree~$S$, both with branch lengths, and the reconstruction~$R$ mapping internal nodes of $S$ to $H$ as
\begin{equation}
P\left(H,S,R|D\right) \propto \int P\left(d_S\right|S) P\left(S|H,R,d_R,\theta\right) P\left(H|d_H\right) P\left(R\right) P\left(\theta\right) \, \dif \theta
\end{equation}
where parameters $\theta = \left(\lambda,\tau,\mu\right)$, the duplication, host-switch, and loss rates, respectively, and are integrated out. The probability of a tree~$T$ given sequence data~$D$ is $P\left(T|D\right) \propto P\left(D|T\right) P\left(T\right)$, where the likelihood is generally calculated with \citeapos{Felsenstein:1981} algorithm or a more complex multi-locus model, e.g. the multispecies coalescent \parencite{Heled:2010a}, and the prior with either a coalescent model \parencite{Kingman:1982} or a birth-death speciation model \parencite{Gernhard:2008}. In this case, the prior on the symbiont tree is represented instead by the reconstruction likelihood $P\left(S|H,R,d_R,\theta\right)$.

@@

\paragraph*{The host-switch operator.}

I implemented the described model in the Java language as a plugin for the BEAST program \parencite{Drummond:2012} for Bayesian evolutionary analysis via \ac{MCMC} simulation. There were several advantages , including  the existing \ac{MCMC} framework and evolutionary library, the number of models, and the ability to utilize any future models developed for BEAST. The source code is available under the GNU General Public License at \url{http://www.github.com/phylocomputing/BECKY}. I provide a Python script to help modify existing BEAST XML analysis specifications to incorporate my cophylogeny model.

\paragraph*{Evaluation on simulated data.}
In addition to the MCMC components

>>>>

Organismic symbioses, or interactions between individuals of two or more different species, constitute a fundamental aspect of many living systems and occur across the biological spectrum. The characteristics of symbiotic relationships can vary greatly, particularly in degrees of cooperation (parasitism vs. mutualism), fidelity (generalized symbionts utilizing multiple species vs. specialized, species-specific symbionts), and obligation (optional vs. vital to survival) \parencites{Oberprieler:2005}. Understanding how these relationships evolve and, in particular, how symbionts affect their partners' evolution---that is, the coevolutionary processes driven by symbiotic interactions---remain important questions in ecology and evolutionary biology.

@@

\textcite{Haffner:1988} were the first to apply phylogenetic methods

@@

I describe a simple cophylogeny model that approximates and implement it in a Bayesian framework.

@@

\begin{figure}
\centering
\pscophylogeny
\caption{An example of a coevolutionary history, where notation corresponds with the four events depicted in \Fref{fig:events}.}
\label{fig:cophylogeny}
\end{figure}

@@

Given data~$D = \left(d_H,d_S,d_R\right)$, the host sequence data, the symbiont sequence data, and the leaf--leaf associations between the host and symbiont trees, respectively, I define the posterior probability of the host tree~$H$ and the symbiont tree~$S$, both with branch lengths, and the reconstruction~$R$ mapping internal nodes of $S$ to $H$ as
\begin{equation}
P\left(H,S,R|D\right) \propto \int P\left(d_S\right|S) P\left(S|H,R,d_R,\theta\right) P\left(H|d_H\right) P\left(R\right) P\left(\theta\right) \, \dif \theta
\label{eq:cophylogenylikelihood}
\end{equation}
where parameters $\theta = \left(\lambda,\tau,\mu\right)$, the duplication, host-switch, and loss rates, respectively, and are integrated out. The probability of a tree~$T$ given sequence data~$D$ is $P\left(T|D\right) \propto P\left(D|T\right) P\left(T\right)$, where the likelihood is generally calculated with \citeapos{Felsenstein:1981} algorithm or a more complex multi-locus model, e.g. the multispecies coalescent \parencite{Heled:2010a}, and the prior with either a coalescent model \parencite{Kingman:1982} or a birth-death speciation model \parencite{Gernhard:2008}. In this case, the prior on the symbiont tree is represented instead by the reconstruction likelihood $P\left(S|H,R,d_R,\theta\right)$.

@@

\paragraph*{Calculating the event probabilities.}

\paragraph*{Implementation in a Bayesian framework.}

I implemented the described model in the Java language as a plugin for the BEAST program \parencite{Drummond:2012} for Bayesian evolutionary analysis via \ac{MCMC} simulation. There were several advantages to integrating my work into BEAST, particularly the existing \ac{MCMC} framework and evolutionary library consisting of numerous models.

To provide support for rate variation over the symbiont phylogeny, . Due to the additional free parameter, I fix the loss rate~$\mu$ at $1$, such that the duplication and host-switch rates $\lambda$ and $\tau$ are normalized against it.

The priors remained to be defined.

A uniform prior is assumed over $R$.

Finally, I define a simple \textsc{host shift operator} to . An internal node $n_S$ from $S$ is selected at random and is assigned a new host node from the set of host lineages existing at time $t = height(n_S)$.\

The source code is available under the GNU General Public License at \url{http://www.github.com/phylocomputing/BECKY}. I provide a Python script to help modify BEAST XML files to i.

\paragraph*{Evaluation of simulated data.}

I simulate the host tree $H$ under the constant-size coalescent \parencite{Kingman:1982} as implemented in BEAST \parencite{Drummmond:2012}

Two simulations
